---
title: 'Volpe R Course: Session 4, Communicating Results'
author: "| Instructors: Don Fisher, Dan Flynn, Jessie Yang \n| Course webpage: http://bit.ly/volpeR\n"
date: "9/26/2017"
output:
  beamer_presentation:
    includes:
      in_header: beamertemp.tex
    incremental: no
  ioslides_presentation:
    incremental: no
  slidy_presentation: default
fontsize: 8pt
colortheme: dolphin
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height = 3.5)
library(ggplot2)
library(gridExtra)
```


## Overview 

### Last session:

- Data exploration
    + Aggregation
    + Merging and Matching

### Today: Communicating results 

- Regular expression review
- Basic graphics
- `ggplot2` package
- Basic analysis


## Regular expressions

Quick digression to review material from last session!

The use of `sub` above is an example of *regular expression*. These are very flexible and form the basis for all search engines. In R, there are number of ways to use regular expressions, with `grep` and `sub` being two of the most common ones. `regexpr` also provides additional options for those familiar with Perl or Python.

```{r, echo = T, results='hide'}
text1 <- c("Testing", "matching ", " and substitutions")

grep("testing", text1, ignore.case = TRUE) # Pattern matching

sub(" +$", "", text1) # find and replace trailing whitespace
grep("^ ", text1) # find elements which start with a whitespace
grep("^[A-Z]", text1) # find elements which start with a capital letter
```

## Homework review 

*We will review example solutions to the homework assignments here*

### Make summary tables using `aggregate` and other functions

### Make a subset using `grep`

### Use `merge`



## Basic graphics

You have already seen the power of `plot()` in examples. We have used it to make scatterplots, but also to make diagnostic plots of linear models. When in doubt, you can always try to `plot()` an object and R will make a guess about what kind of plot you want.

Example:
```{r}
data(iris) # see ?iris

# Formula (y ~ x) versus (x, y) specification
plot(Sepal.Length ~ Petal.Length, data = iris)
plot(iris$Petal.Length, iris$Sepal.Length) # identical
# plot(Petal.Length, Sepal.Length, data = iris) # fails
```
## Basic graphics (plus some analysis)

You frequently want to analyze what the trend of the data is statistically, not just graphically. We can use a simple linear model and plot the result using `lm()`:

```{r, echo=TRUE, message = F, results='hide'}
plot(Sepal.Length  ~ Petal.Length, data = iris)

summary(linmod <- lm(Sepal.Length ~ Petal.Length, data = iris))

abline(linmod)
```

### Interpretation: 
What is the slope of this line? How would you interpret it?

## Basic graphics (plus some analysis)

We have often looked at boxplots for categorical data. The analysis that corresponds to a simple boxplot is an "Analysis of Variance" or ANOVA. In `R`, you can do this analysis with the `aov` function:

```{r}

plot(Petal.Length ~ Species,
     data = iris)
```
```{r, results = 'hide'}
summary(aovmod <- aov(Petal.Length ~ Species, data = iris))
coef(aovmod)
```

### Interpretation:
What is the model estimate for the mean value of *Iris versicolor* petal lengths? 

## Basic graphics
Back to graphics! The power of the base graphics package in R is control of each element of the plot. The challenge is knowing how to exercise that control.

`par()` sets graphics parameters and contains much of what you want to do. Highlights from `par`:


    mfrow - matrix format. Set to c(1, 2) for two panel figure, adjacent 
    pch   - plotting character. 
    lty   - line type. Try lty = 2 or lty = 3
    cex   - character expansion. Very useful for making symbols more visible
    mar   - margins. Set to mar = rep(1, 4) for narrow margins all around.
    col   - colors, see colors() for full list or
    www.nceas.ucsb.edu/~frazier/RSpatialGuides/colorPaletteCheatsheet.pdf)

`?par` for the full list.

## Basic graphics

```{r, fig.height = 4.25}
par(mfrow = c(1, 2), 
    pch = 16,
    cex = 1.5) # makes everything bigger

plot(Sepal.Length ~ Petal.Length, 
     col = "grey80",
          data = iris)

plot(Sepal.Width ~ Petal.Width, 
     col = "dodgerblue",
     pch = 15, # can override par settings
          data = iris)
```


## Basic graphics
`barplot()` can be used to create a bar chart, but it is a bit clunky to add error bars. Basic workflow:

```{r, fig.height = 4.25}
means <- tapply(iris$Sepal.Length, iris$Species, mean)
sds <-  tapply(iris$Sepal.Length, iris$Species, sd)

# save the location of the bars as 'bp'
bp <- barplot(means)

arrows(bp, means+sds, # see ?arrows
       bp, means-sds, 
       angle = 90,
       code = 3)
```




## Basic graphics
Improve a graphic by ensuring that there are clear labels (`arrows` code not shown).
<!-- two versions, one with arrows code not shown for conciseness -->
```{r, eval=FALSE, include=TRUE, echo = TRUE}
par(mar = c(3, 4, 3, 1)) # Narrowing margins
bp <- barplot(means, 
              ylab = "Mean sepal length (cm, ? 1 s.d.)",
              main = "Iris species sepal lengths",
              xaxt = "n", # suppress axis plotting here
              ylim = c(0, 7.5)) 

axis(side = 1, at = bp, 
     labels = paste("Iris", c("setosa", "versicolor","virginica")),
     font = 3)
```
```{r, echo=FALSE}
par(mar = c(3, 4, 3, 1)) # Narrowing margins
bp <- barplot(means, 
              ylab = "Mean sepal length (cm, ± 1 s.d.)",
              main = "Iris species sepal lengths",
              xaxt = "n",
              ylim = c(0, 7.5))
arrows(bp, means+sds, bp, means-sds, 
       len = 0.1, 
       angle = 90,
       code = 3)
axis(side = 1, at = bp, 
     labels = paste("Iris", c("setosa", "versicolor","virginica")),
     font = 3)
```

## Basic graphics

Save your graphics typically as pdf or jpeg. You can do this by putting `pdf()` and `dev.off()` around your code, or by using `dev.print()`.

```{r, message=FALSE, warning=FALSE}
pdf("Test graphic.pdf", width = 4, height = 6)
plot(1:10, pch = 1:10, col = 1:10, cex = 2)
dev.off()

plot(1:10, pch = 11:20, col = rainbow(10), cex = 2)
dev.print(device = jpeg, file = "Test graphic.jpg", height = 500, width = 800)

```

## `ggplot` graphics

A different model of graphics has recently become popular. `ggplots` use a different 'grammar' of graphics, and can be great for quickly making visually-appealing graphics. Mastering the control of each element can take time.

```{r}
library(ggplot2) # install.packages('ggplot2')
#       Dataset               x               y         type of plot
gp1 <- ggplot(iris) + aes(Sepal.Length, Petal.Length) + geom_point()

gp1 

```

## ggplot graphics

You can use one basic graph data set and do different versions of it. Two ways of visualizing a factor grouping:
<!-- show but do not evaluate -->
```{r, eval = FALSE}
gp1 + geom_point(aes(color = Species))
gp1 + facet_wrap(~ Species)
```
<!-- evaluatate but do not show -->
```{r, echo = FALSE, fig.height = 4.5}
ex1 <- gp1 + geom_point(aes(color = Species))
ex2 <- gp1 + facet_wrap(~ Species)
grid.arrange(ex1, ex2, ncol =2)
```

## Interactive 

These can be made interactive in an HTML document using `plot.ly`. See https://plot.ly/r/ if interested in trying this! You will need to download several packages, so this code will not work for you immediately:

<!-- change eval = T for HTML version --> 
```{r, warning=F, message=F, eval = F}
library(plotly)
ggplotly(ex1)
```

## Tables in R

Carrying out any analysis in R generates a lot of tables. A few options:

- Copy and paste
- Sink
- Save to text file

Simplest is copy and paste from the console window into Excel, then use text-to-columns. But not repeatable!

## Tables in R

You can 'sink' your output to a text file. First, set up all your analysis, then put `sink()` before and after your block of code.

```{r, eval = F}
sink("Test Sink.txt")
with(iris, cor.test(Sepal.Width, Petal.Width))
table(iris$Species)
sink()
```

Or, more commonly, output a data frame to a text file directly:
```{r}
write.csv(iris[1:10,], file = "Iris Test.csv", row.names = F)
```

## R Markdown

A better way for writing a report is to take advantage of the relatively recent tool, R Markdown. A few resources:

http://rmarkdown.rstudio.com/
http://rmarkdown.rstudio.com/authoring_quick_tour.html

Go to File > New File ... > R Markdown

Select HTML or Word as the default output, and then try clicking 'knit'.

You may need to `install.packages("rmarkdown")`


## Homework review: Outliers

*Time permitting, we will go back to the Hubway data and examine the question of outliers*

We will review `tripduration` outliers from the Hubway data.


