---
title: 'Volpe R Course: Session 3, Data Exploration'
author: "| Instructors: Don Fisher, Dan Flynn, Jessie Yang \n| Course webpage: http://bit.ly/volpeR\n"
date: "9/19/2017"
output:
  slidy_presentation: default
  ioslides_presentation:
    incremental: no
  beamer_presentation:
    includes:
      in_header: beamertemp.tex
    incremental: no
fontsize: 8pt
colortheme: dolphin
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height = 3.5)
library(ggplot2)
library(gridExtra)
```


## Overview 

### Last session:

- Writing functions
- Writing loops


### Today: Data exploration

- Aggregating data
- Matching
- Reshaping data
- Basic data analysis
- Saving work and sharing

## Homework

*We will review your work these steps at the end of the session, time permitting*

### Loops

1. R skill: write two loops. One should loop generate random subsets of your data; Another should loop over rows in a data frame and perform some operation

### Functions


## Data manipulation

Hubway `tripduration` outlier test: 
Read in the Hubway data from the course page and look at tripduration. How would you decide which data should be considered outliers?

## Aggregating data


## Aggregating data

Three functions are most useful for aggregating data: `aggregate`, `tapply`, and `apply`.
Using the merged file, create mean net fuel consumption tables:

```{r, results = 'hide'}
# Two index vectors, using aggregate
aggregate(fuel.temp$NetFuel, 
          by = list(fuel.temp$Region, fuel.temp$State), FUN = mean)

# Or with only one index vector, using tapply
tapply(fuel.temp$NetFuel, fuel.temp$Region, mean)
```
Use apply to look across multiple columns or rows

```{r, results = 'hide'}
apply(fuel.temp[3:4],     # data frame to look at (should be two-dimensional)
      2,                  # 1 is for rows, 2 is for columns
      mean)               # Function
 
```

### Exercise: 
Create a table of *annual* net fuel consumption for each region. 



## Matching and merging


## Regular expressions

The use of `sub` above is an example of *regular expression*. These are very flexible and form the basis for all search engines. In R, there are number of ways to use regular expressions, with `grep` and `sub` being two of the most common ones. `regexpr` also provides additional options for those familiar with Perl or Python.

```{r}
text1 <- c("Testing", "matching ", " and subustitutions")
sub(" +$", "", text1) # find and replace trailing whitespace
grep("^ ", text1) # find elements which start with a whitespace
grep("^[A-Z]", text1) # find elements which start with a capital letter
```

## Merging data frames

Merging data between different sources requires having a common column name or row name. The two functions most often used are `merge` and `match`. Type the following:

```{r, results = 'hide'}
example(merge)
authors
books

merge(authors, books, by.x = "surname", by.y = "name") 
# and do it again again, adding `all.y = TRUE`
```

We can use match to carry out a similar operation. Match is more flexible than merge, and therefore more difficult to use!

```{r, results='hide'}
m1 <- match(authors$surname, books$name) # returns index for the second element which match the first

books[m1,]

newdata <- data.frame(authors, books[m1,])
```


## Merging data frames

`match` can also be used in many other contexts, besides merging data frames. Additional functions for matching to know are `which`, `grep`, and `intersect`.

```{r}
which(books$name == "Ripley")
grep("Rip", books$name) # partial string matching
intersect(books$name, authors$surname) 
# unique shared elements, order unimportant.

```



## Saving your work

After doing all this work, you will want to save the results. You can save R objects in your current workspace to file type called `.RData`. Then you can load these R objects back in at a later session, without having to carry out this preparation work.

For example, to save all objects in the current workspace:

`save(list = ls(), file = "My_Prepped_Data.RData")`

and load it back in using

`load("My_Prepped_Data.RData")`

To save specific R objects, you will name them in the list as strings, such as

`save(list = c('fuel.temp', 'avg.temp'), file = "My_Prepped_Data.RData")`

You can also consider making a script that does all this preparation work, and save it as a file called something like `Analysis Prep.R`. Then, in your data analysis script or `.Rmd` file, you can automatically run that script by inserting the command:
`source('Analysis Prep.R')`.

<!-- Time permitting, review homework:

## Homework review: Loops
Loop to generate random subsets of your data:

```{r}
# Get 10 samples of data from each species:
sampx <- vector()
for(i in levels(iris$Species)){
  iris.sp.i <- iris[iris$Species == i,]
  sampx <- rbind(sampx, iris.sp.i[sample(rownames(iris.sp.i), 10),])
  }
```

Loop over rows in a data frame and perform some operation:
```{r}
# Subtract mean from each 
means <- apply(iris[,1:4], 2, mean)
newdat <- vector()
for(i in 1:nrow(iris)){
  newdat.x <- iris[i,1:4] - means
  newdat <- rbind(newdat, data.frame(newdat.x, iris[i,5]))
  }
# Examine it: hist(newdat[,2]); hist(iris[,2])
```


-->


